---
title: "Regression Models Course Project - Kevin O'Leary"
output:
  pdf_document:
    fig_crop: no
    fig_height: 3.55
    fig_width: 6
geometry: tmargin=1cm, lmargin=1cm, rmargin=1cm, bmargin = 1.5cm
fontsize: 5pt
---

# Introduction

To get a quick glimpse of the answer, we can compute the mean of mpg by transmission and we find that manual has higher corresponding mpg on average than automatic transmission (Figure 1.).



```{r echo=FALSE}
aggregate(mpg~am, data=cars, mean)

summary(lm(mpg ~ am, data = cars))$coef
```

```{r}
vif(minmod)
```

It appears that cyl is a little highly correlated. Let's test before we throw it out.

```{r echo=FALSE}
minmod2 = update(minmod,.~. - cyl)

vif(minmod2)
```
This looks better, but let's do an ANOVA test to see how significant the change is.

```{r echo=FALSE}
anova(minmod2, minmod)
```

Not very significant but I won't disregard it without seeing what it does to the adjusted R squared value.
```{r echo=FALSE}

summary(minmod)$adj.r.squared
summary(minmod2)$adj.r.squared
```

On balance, I think removing it would be too costly to the model fit so it's staying.



